Increased productivity using Atomify - a real-time LAMMPS visualizer

### 2018-06-03 ms.tex
Wrote more intro

\documentclass[aps,pre,twocolumn,letterpaper,floatfix,nofootinbib]{revtex4}
\usepackage{graphicx} 
\usepackage{amsmath,amssymb,amsfonts} 
\usepackage{mathtools}
\usepackage{pdfpages}
\usepackage{afterpage}
\usepackage[hidelinks]{hyperref} 
\usepackage{epstopdf}
\usepackage{todo}
\usepackage{menukeys}
\usepackage{siunitx}
\begin{document}

@@

\section{Introduction}
Our nature is a very complex system with complicated rules that determine how and where mass, energy and momentum is transferred.
Many macroscopic observable effects result from microscopic details on the atomic level such as melting of ice, and even life.
Understanding the world from a bottom-up perspective requires us to be able to model a physical system from the atomic scale where the timescale is of order femtoseconds.
Many physical and chemical processes happen on nanosecond timescales, some even during a few picoseconds, which means thousands or millions of timesteps of \SI{1}{\femto\second} need to be calculated.
Much has happened since the first molecular dynamics simulation of hard spheres in the 1950s\citep{alder1957phase,alder1959studies}, and later, in 1964, a more realistic system with the Lennard Jones potential studying diffusion in argon\citep{rahman1964correlations}.
The computing power has seen a tremendous increase from the CDC 3600 used in \citep{rahman1964correlations}, which performed less than 1 million instructions per second whereas modern 3 Ghz processors in typical laptops can do about 100 billion instructions per seconds.
With this rapidly increase of computing power, the lengthscales and timescales achievable in numeriments\footnote{Numeriment is a numerical experiment, or a simulation on the computer.} is finally catching up with some experiments (ref).

@@

The work in this paper is focused on LAMMPS.
Although the software is well-documented\footnote{The documentation of LAMMPS is known for being very detailed precise.} and enables the user to perform
advanced simulations, it can be demanding to learn how to use it well.
LAMMPS has a huge list of implemented features which makes it very flexible with its own scripting language.
A LAMMPS simulation is performed by a series of commands which are executed one by one.

Physical quantities such as energy, temperature and stress can be measured by adding \textit{computes}.
The time evolution of these quantities can in turn be saved to file for plotting
and further analysis.
Particle trajectories can be saved to a file and visualized by tools such as
VMD\citep{Humphrey1996Vmd} or OVITO\citep{Stukowski2009Visualization}.
The lack of an intuitive GUI and visualization\footnote{GUI is listed as a non-feature on the web page:
\url{http://lammps.sandia.gov/non_features.html}.} makes the workflow somewhat
complicated, see figure \ref{fig:flowchart}.

Atomify combines script editing, simulation,
visualization, and analysis in one single application.
It uses LAMMPS as a library and synchronizes after each timestep,
copying the relevant information from LAMMPS to the graphics processing unit
(GPU) for visualization and data plotting.
The visualization is optimized for high performance by raytracing billboards.
This allows visualization of millions of atoms with good framerate on desktop
computers. Atomify is designed with simplicity as main goal.

A simplified version of Atomify is made available for mobile platforms (Android
and iOS).
This version of Atomify provides access to some of the examples and is intended
for exploration of molecular dynamics simulations by students.
It also works as a demonstration of the visualization capabilities of the
desktop version of the software.
The authors are optimistic that teachers can use the visualizations provided by
Atomify to explain microscopic interactions underlying concepts in topics such
as thermodynamics, geophysics, and biology.
\begin{figure*}
	\centering
	\includegraphics[width=0.95\textwidth]{gui.png}
	\caption{%
    Overview of the Graphical User Interface (GUI) in Atomify.
    The toolbar (1) is used to change between the different modes, such as 
    editing, viewing and analyzing the running simulation, 
    in addition to playback controls.
    The file browser (2) contain the list of open LAMMPS scripts or data files.
    The text editor (3) provides code highlighting.
    The viewport (4) shows the current simulation.
    The simulation properties (5) shows the simulations details, gives access to 
    LAMMPS objects such as groups, regions and computes, and the rendering
    properties (6) lets you change the viewport settings, such as lighting and
    draw mode.
    TODO: Add numbers to figure.
    }
	\label{fig:gui}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[width=0.95\textwidth]{flowchart.png}
	\caption{The typical workflow when using LAMMPS consists of multiple steps using several applications. a) A LAMMPS script is written using a text editor. b) Simulation is performed from the terminal. c) and d) visualization and analysis is performed using Ovito and \url{http://bit.ly/logplotter}. Atomify simplifies the workflow by combining all steps into a single application. }
	\label{fig:flowchart}
\end{figure*}

\section{Features}
The main features of Atomify are script editing,
live visualization, and live plotting of computed quantities.
In addition, Atomify comes packaged with multiple examples,
many of which have been made by members of the LAMMPS community.
These examples are easily explored within Atomify.

Atomify provides a lightweight script editor with syntax highlighting and line
numbers.
% TODO add built-in help and autocompletion to Atomify and mention here
Multiple files can be open at the same time and the state persist across working sessions.
The script can easily be restarted by clicking the button in the GUI or using
the default shortcut (\keys{Ctrl+R} in Windows and Linux or \keys{\cmd+R} in macOS).
This enables you to quickly do changes in the script and see the changes immediately.

Certain Atomify-commands can be added to the comments of a regular LAMMPS script
to modify properties of the GUI.
All these start with \keys{\texttt{\#}} so that LAMMPS ignores them as they are
interpreted as comments.
For instance, the color and size of the atoms can be set based on the type by
the command \keys{\texttt{\#/atom}}.
To set the color to white and radius to 2.0 for particle type 1,
the user can apply the command \keys{\texttt{\#/atom 1 2.0 \#ffffff}}.
Certain built-in atom types have default values and can be accessed with
\keys{\texttt{\#/atom 1 oxygen}} for simplicity.
The initial camera position can also be set similarily.
For a full list of the available commands, see the Atomify
documentation\citep{atomifydocumentation}.

Atomify provides live visualization of particles in a simulation in its 3D view. 
The particles can be colored based on any produced per-atom quantity in LAMMPS such as
atom type, velocity, force or a variable formula.
The coloring scheme can be changed while the simulation is running by hovering
the mouse cursor over groups in the dashboard.
Light colors and strength can also be changed in the dashboard.
This will be described in further detail in the following section.

It is possible to organize particles in groups and regions in LAMMPS.
These are shown in the dashboard in Atomify, which also provides controls to
highlight or hide specific groups or regions.
To highlight a group or region, the user only needs to hover the name in the
dashboard.
The visibility icon can be clicked to hide a group or region.

Parameters for rendering, such as the light strength,
attenuation, and color can be modified in the Rendering tab of the dashboard.
See the above section about the script editor for information about how to
modify the color of the particles.

In LAMMPS, computes are computations performed on a group of atoms.
Variables are user defined expressions which can be used to combine values in
calculations.
Once defined in a LAMMPS script, these can be plotted as a function of time in
Atomify.
This is done by clicking the name of the variable or compute in the simulation
tab.
We will explore plots in further detail in the following section.

Any operation that is applied to the system in a LAMMPS timestep or minimization
step is called a ``fix''.
These fixes are defined in the LAMMPS script using the ``fix'' keyword.
Atomify registers all defined fixes and lists these in the dashboard.

In LAMMPS, computes are physical quantities calculated based on the running
simulation, such as energy, temperature and stress.
% TODO better word for measures?
These are typically macroscopic measures derived from the microscopic
interactions between the atoms.
Computes and variables are shown in the Atomify dashboard,
along with information about the regions, groups and fixes as shown in figure
\ref{fig:dashboard}.

\begin{figure}
	\centering
	\includegraphics[width=0.3\textwidth]{dashboard.png}
	\caption{Dashboard}
	\label{fig:dashboard}
\end{figure}

Computes can be plotted as a function of time and visualized while the 
simulation is running.
Quantities that are measured per atom can be plotted in a histogram or a 2D
average for each region.
Atomify provides live visualization of the compute values using various plotting
techniques.
For instance, the temperature of the simulated system can be plotted as a
function of time as seen in figure \ref{fig:temperature}.

% \begin{figure}
% 	\centering
% 	\includegraphics[width=0.45\textwidth]{figures/temperature.png}
% 	\caption{Atomify can visualize computes defined in LAMMPS. 
%     Here the temperature is plotted as a function of time.}
% 	\label{fig:temperature}
% \end{figure}

\section{Implementation}

@@

\section{Conclusion}
%
We have presented Atomify, a software for live visualization of molecular
dynamics simulations in LAMMPS.
Atomify simplifies the workflow of researchers by combining script development,
analysis and visualization in an intuitive user interface.
The software can also be used to improve training of students using LAMMPS by
providing immediate, visual feedback to script changes.

@@

\subsection{Future work}

>>>>

\documentclass[aps,pre,twocolumn,letterpaper,floatfix,nofootinbib]{revtex4}
	% for \code color
\usepackage[dvipsnames]{xcolor}
\usepackage{graphicx} 
\usepackage{amsmath,amssymb,amsfonts} 
\usepackage{mathtools}
\usepackage{pdfpages}
\usepackage{afterpage}
\usepackage[hidelinks]{hyperref} 
\usepackage{epstopdf}
\usepackage{todo}
\usepackage{menukeys}
\usepackage{siunitx}
\usepackage{listings}
% Pretty formating of c++ (see \cpp command)
\usepackage{xspace}
\newcommand*{\cpp}{C\ensuremath{++}\xspace}

\lstdefinelanguage{LAMMPS}{
    morekeywords={units,atom_style,lattice,region,create_box,create_atoms,pair_style,pair_coeff,neigh_modify,mass,velocity,fix,run},
    sensitive=false, % keywords are not case-sensitive
    morecomment=[l]{//}, % l is for line comment
    morecomment=[s]{/*}{*/}, % s is for start and end delimiter
    morestring=[b]" % defines that strings are enclosed in double quotes
}

\lstset{
numbers=left, 
numberstyle=\small, 
numbersep=8pt, 
frame = single, 
language=LAMMPS, 
framexleftmargin=15pt,
xleftmargin=0.65cm,
xrightmargin=0.1cm}

% Highlighte inline code
\definecolor{light-gray}{gray}{0.95}
%\newcommand{\code}[1]{\mbox {\texttt{#1}}}
\newcommand{\code}[1]{\colorbox{light-gray}{\color{RawSienna}\texttt{#1}}}

\begin{document}

@@

\section{Introduction}
Our nature is a very complex system with complicated rules that determine how, and where mass, energy and momentum is transferred.
Many macroscopic observable effects result from microscopic details on the atomic level such as melting of ice, and even life.
Understanding the world from a bottom-up perspective requires us to be able to model a physical system from the atomic scale where the timescale is of order femtoseconds.
Many physical and chemical processes happen on nanosecond timescales, some even during a few picoseconds, which means thousands or millions of timesteps of \SI{1}{\femto\second} need to be calculated.
Much has happened since the first molecular dynamics simulation of hard spheres in the 1950s\citep{alder1957phase,alder1959studies}, and later, in 1964, a more realistic system with the Lennard Jones potential studying diffusion in argon\citep{rahman1964correlations}.
The computing power has seen a tremendous increase from the CDC 3600 used in \citep{rahman1964correlations}, which performed less than 1 million instructions per second whereas modern 3 Ghz processors in typical laptops can do about 100 billion instructions per seconds.
With this rapidly increase of computing power, the lengthscales and timescales achievable in numeriments\footnote{Numeriment is a numerical experiment, or a simulation on the computer.} is finally catching up with some experiments (ref).

@@

\lstinputlisting[language=LAMMPS,caption={A minimal LAMMPS input script that simulates a Lennard Jones solid at reduced temperature $T^* = 1.0$ with a reduced density of $\rho^* = 0.8442$.},label={lst:simple.in}]{simple.in}

The work in this paper is focused on LAMMPS.
Although the software is well-documented\footnote{The LAMMPS documentation is known for being very detailed and precise.} and enables the user to perform advanced simulations, it can be demanding to learn how to use it well.
The core principles of molecular dynamics is so simple that high school students can easily understand the concepts, but even undergraduate students may have problems with the terminology of statistical mechanics which is used in several of the commands.
For instance, to simulate an \textit{NVE} simulation (constant number of particles $N$, volume $V$ and energy $E$), we use the \code{fix nve} command.

A LAMMPS simulation is performed by a series of commands, which are executed one by one.
LAMMPS has a huge list of implemented features which makes it very flexible, especially with its own, powerful scripting language.
In listing \ref{lst:simple.in}, we show an example of a minimalistic LAMMPS script that simulates 4000 Lennard Jones atoms in an FCC lattice with reduced temperature $T^*=1.0$ and reduced density $\rho^* = 0.8442$.
This simulation does not print much thermodynamic data, but it creates a file \code{trajectory.bin} which containes the trajectory of all atoms at every timestep.
Other programs such as VMD\citep{Humphrey1996Vmd} or OVITO\citep{Stukowski2009Visualization} can be used to visualize the file as a postprocessing step.

If we want to measure physical quantities such as energy, temperature and stress, they can be accessed through a series of commands called \textit{computes}.
For instance, the commands \code{compute ke} and \code{compute pe} measures the kinetic and the potential energy.
Using \textit{fixes}, we can create time averages and save these quantities to one or more files for plotting and further analysis.
We also often put atoms in different \textit{groups}, either by their position being in a \textit{region}, or their atom type or more advanced criteria.
Performing different operations on different groups, or just measuring physical quantities on a specific group is then easily done since almost all commands have \code{group} as an input parameter.

\begin{figure*}
	\centering
	\includegraphics[width=0.95\textwidth]{flowchart.png}
	\caption{The typical workflow when using LAMMPS consists of multiple steps using several applications. a) A LAMMPS script is written using a text editor. b) Simulation is performed from the terminal. c) and d) visualization and analysis is performed using Ovito and \url{http://bit.ly/logplotter}. Atomify simplifies the workflow by combining all steps into a single application.}
	\label{fig:flowchart}
\end{figure*}

The lack of an intuitive GUI and visualization\footnote{GUI is listed as a non-feature on the LAMMPS web page: \url{http://lammps.sandia.gov/non_features.html}.}
makes the workflow complicated, see figure \ref{fig:flowchart}.
A simulation with full analysis usually involves writing the script in a text editor, running the simulation using the terminal before analysis is done with a visualization program and additional scripts written in MATLAB or Python.
When creating a new simulation, the initial state (system box, positions of all particles, groups and regions) can be tricky to set up right, especially since groups and regions are not automatically written to the output file through the \code{dump} command.
The time from an idea to a visual feedback can then be destructive for the creativity where immediate feedback would be ideal.

We have solved this problem by creating Atomify, a real-time LAMMPS visualizer that combines script editing, simulation, visualization, and analysis in one single application.
It uses LAMMPS as a library and hooks into the LAMMPS simulation cycle so synchronization of atom positions and physical quantities can be performed as often as every timestep.
Most features of LAMMPS are automatically supported, except a few due to technical reasons discussed in section \ref{sec:future}.

The visualization is performed using the powerful Qt3D library where advanced rendering techniques such as billboard raytracing is used to achieve maximum performance.
This allows visualization of millions of atoms with good framerate on modern desktop computers.
Atomify is designed with simplicity as its main goal.

Since we have used the Qt framework to develop Atomify, and LAMMPS is written in object oriented \cpp, a simplified version is also made available on mobile platforms (Android and iOS).
This version of Atomify provides access to some of the example simulations and is intended for exploration of molecular dynamics simulations by students and scientific outreach.
It also works as a demonstration of the visualization capabilities of the desktop version of the software.
The authors are optimistic that teachers can use the visualizations provided by
Atomify to explain microscopic interactions underlying concepts in topics such
as thermodynamics, geophysics, and biology.

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{gui.png}
	\caption{%
    Overview of the Graphical User Interface (GUI) in Atomify.
    The toolbar (1) is used to change between the different modes, such as 
    editing, viewing and analyzing the running simulation, 
    in addition to playback controls.
    The file browser (2) contain the list of open LAMMPS scripts or data files.
    The text editor (3) provides code highlighting.
    The viewport (4) shows the current simulation.
    The simulation properties (5) shows the simulations details, gives access to 
    LAMMPS objects such as groups, regions and computes, and the rendering
    properties (6) lets you change the viewport settings, such as lighting and
    draw mode.
    TODO: Add numbers to figure.
    }
	\label{fig:gui}
\end{figure*}

In this paper we will go through a case study, an example simulation, where we
use many of the useful features Atomify provides.
We start by discussing the GUI in section \ref{sec:features} where the main features are summarized.
In section \ref{sec:casestudy}, the case study is introduced where we go through a typical example
of a simulation that uses regions and groups to perform different integration on different atoms.
Details about how Atomify runs LAMMPS is found in section \ref{sec:implementation} where the data flow and the threading model is explained.
Finally, we summarize the work in our paper in section \ref{sec:conclusion} with some thoughts about future outlooks in section \ref{sec:future}.

\section{\label{sec:features}Summary of features}
The main features of Atomify is summarized as script editing,
real-time visualization and plotting of physical quantities.
In addition, Atomify comes with multiple examples,
many of which have been made by members of the LAMMPS community.
It provides a lightweight script editor with syntax highlighting and line numbers.

Multiple files can be open at the same time and the state persist across working sessions.
The script can easily be restarted by clicking the button in the GUI or using
the default shortcut (\keys{Ctrl+R} in Windows and Linux or \keys{\cmd+R} in macOS).
This enables the user to get immediate visual feedback on changes in the script.

\section{\label{sec:casestudy}Case study: flow through nanoporous media}
Fluid flow is often characterized with permeability, the inverse flow resistance of a material.
The permeability can be measured using Darcy's law when applying a pressure gradient.

Humans use an enormous amount of energy, most of which comes from chemical reactons in burning hydrocarbons.
\todo{Short motivation about flow in nanoporous media}

\section{\label{sec:implementation}Implementation}

@@

\section{\label{sec:conclusion}Conclusion}
%
We have presented Atomify, a software for live visualization of molecular
dynamics simulations in LAMMPS.
Atomify simplifies the workflow of researchers by combining script development,
analysis and visualization in an intuitive user interface.
The software can also be used to improve training of students using LAMMPS by
providing immediate, visual feedback to script changes.

@@

\section{\label{sec:future}Future work}
